---
title: "Practical 2 - Maftools"
author: "Conor O'Donoghue"
date: "March 5, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Install the Bioconductor package maftools

```{r install maftools, eval=FALSE}
BiocManager::install("maftools")
```

```{r load maftools, message=FALSE}
library(maftools)
```


# 2. Read the maf format somatic variants into a maftools object.

```{r read.maf}
laml = read.maf('somatic.mafplus')
laml
```

# 3. 
## a) How many mutations are there in this dataset?
According to the summary table, the total of all mutations is 760.

## b) How many genes are affected?
Again according to the summary table, 483 genes were affected.


# 4. 
## a) Produce a mafSummary plot
```{r plotmafSummary}
plotmafSummary(maf = laml, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE)
```


## b) What is the most mutated gene?
According to the plot, MUC4 is the most mutated gene, with just under 30 mutations.


# 5. Investigate whether there is any known association between the gene in step 4 and pancreatic cancer.
According to genecards (https://www.genecards.org/cgi-bin/carddisp.pl?gene=MUC4), 
MUC4 is a gene that codes for an integral membrane glycoprotein, and thus has function in cell adhesion. The UniProtKB summary of the gene states that it may play a role in tumor progression, and promotes tumor growth mainly due to repression of apoptosis. The page includes mRNA expression of MUC4 by UniProt/SwissProt, which found that it was minimally or not expressed in normal pancreas or chronic pancreatitis, but was found to be highly expressed in pancreatic tumors and pancreatic tumor cell lines.

# 6. Produce a lollipop plot for MUC4 in the pancreatic samples.
```{r lollipopPlot}
lollipopPlot(maf = laml, gene = 'MUC4', AACol = 'AAChange', showMutationRate = TRUE)
```

# 7.
## a) Produce a plot of mutation types using plotTiTv.
```{r plotTiTv}
laml.titv = titv(maf = laml, plot = FALSE, useSyn = TRUE)
#plot titv summary
plotTiTv(res = laml.titv)
```


## b) Which mutation type is  most common in the pancreatic samples?
Transitions were found to be more common than transversions, with T->C being the most common mutation type.


# 8. How does the mutation rate per Mb in the Texas samples compare to the rate in the TCGA cohorts? Produce a plot to illustrate this comparison.
```{r tcgaCompare}
laml.mutload = tcgaCompare(maf = laml, cohortName = 'Texas', logscale = TRUE, capture_size = 50)
```
The mutation rate appears to be around the middle compared to the TCGA cohorts -- it's not among the highest, but not among the lowest either.

# 9. Are there any mutations that tend to co-occur/tend not to co-occur in the pancreatic cancer samples? Produce a plot to illustrate these.
```{r somaticInteractions}
somaticInteractions(maf = laml, top = 25, pvalue = c(0.05, 0.1))
```
It appears that KRTAP4-5 and KRTAP4-11 have the most significant rate of co-occurance.
KCNN3 and ATXN1 as well as PAK2 with both FCGBP and HRNR also reached a significant p-value for co-occurance.

KCNN3 and MUC4 as well as LSR and NBPF10 have a p-value below 0.1 for mutual exclusivity, but neither reach p<0.05, 
so we ought not to conclude that any pair of genes have been shown to be mutually exclusive.

# 10. Are any of the actionable variants in OnkoKB found in any of the Texas samples?

```{r}
# Can use subsetMaf and provide genes from oncokb dataframe
oncokb = read.csv('cancerGeneList.tsv', sep='\t')
laml_subset <- subsetMaf(maf=laml, genes=oncokb$Hugo.Symbol)
laml_subset
```

According to the new maf object, there are 31 genes across 9 samples that would be actionable according to OnkoKB. We would now be able to use the laml_subset object to obtain any additional information about the expression of these genes in our samples.